Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-26 analysis/paper/paper.Rmd
working on methods section

# Materials and archaeological background

Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from the largest excavation section at Kiwulan, northeastern Taiwan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)). Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides . Yilan is an ideal context to study peripheral colonial influence because it was relatively isolated by natural barrier that reduced frequency of direct contact with the Europeans presences, the Spanish and the Dutch settled in northern Taiwan. The chronology of Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007]. This paper focuses on the upper component dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period defined by the European presence in Taiwan in the early 17th century. The Dutch firstly came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626. In 1642, the Spanish was expelled by the Dutch, who took over their forts at Helping dau in Keelung, and in Tamsui. Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

The upper component experienced foreign contacts including the Europeans in the 17th century and great waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stonewares, and ornaments elements such as beads were trade goods commonly found in the upper component that indicates the frequent long-distance trade activities in the 17th century during the European presence in northern Taiwan. In addition to artifacts, features such as burials, middens, and post-holes with in-situ posts widespread across 1-2 m thick deposit explain that Kiwulan was a continuously occupied and large settlement site [@Chen2007]. To compare different foreign influences, the upper component were classified into three phases, the pre-European, European, and Chinese periods. The principle of identification of three phases is based on 32 radiocarbon ages, excavation depth, consistency of contexts, and types of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars prevailing in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011]. The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.

The predominant artefact found is locally made ceramics, which distributed throughout the sequence across the site. More than 550 thousands of sherds were excavated, and around 1,200 vessels were reconstructed. It is interesting to note that there are only two forms of local vessels, one is cooking pot and another one is steamer made of two cooking pots stacked together with a filter clay layer in between. Those pots show high consistency in shape. They are globular body with short neck and wide mouth. The exterior surface below the neck is treated with a wide variety of impressed geometric motifs. They were believed used for cooking based on the evidence of charred residues and carbon deposits commonly observed on the interior and soot on the exterior. Moreover, a lack of evidence of other utilitarian earthenware ceramics suggests this globular pot was mainly used for daily cooking. They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand. In addition to Kiwulan site, this kind of pot was widely found at other archaeological sites during the same period in Yilan Plain.

@@

Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate. The most common type of sherds, also the dominant one, has a high percentage of inclusion (25-40%), which is composed mainly by argillite (15-40%), followed by metasandstone (1-10%), sandstone (1-6%), and quartz (1-5%). The size of the particles ranges from 500 micron to 1300 micron (=1 mm). In general, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz. The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016]. It seems there is no obvious changes in the inclusions over time.

@@

A total of `r total_no` relativity intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for Ellipic Fourier analysis. Most of them cover every components from rim to bottom. Given that the decline of indigenous population in the Chinese period might have impact on the interpretation, we examined the spatial distribution of pots analyzed in this paper. The pottery distributed randomly across the sampling area without any clusters could rule out the bias caused by the population dynamics (Figure \@ref(fig:pottery-spatial-pattern)). To fully explore the pottery standardization, we also take the metric measurements into account for comparison. The total of 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) were measured. The metric attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body. For each pot, two measurements for thickness and diameter were took to compute an average value. To obtain information about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.

```{r pottery-spatial-pattern, fig.cap= "The spatial distribution of the pottery selected for shape analysis. The quantity is indicated by the color scale."}
library(sf)
library(viridis)

# read in the spatial data
AD_zone <- invisible(st_read(here("analysis", "data", "raw_data", "AD_zone.shp"), quiet = TRUE))
units_40 <- invisible(read.csv(here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv")))

AD_zone_tidy <-
  AD_zone %>% 
  mutate(Pit =paste0("P0",Id)) %>% 
  inner_join(units_40, by = c("Pit" = "the_sq")) %>% 
  select(Pit, geometry)

# tidy pottery data
pottery_count <-
  pottery_data %>% 
  select(Site, Context, Pit, Layer, No., phase) %>% 
  count(Pit, phase)

# join two dataset to get coordinate
pottery_count_join <-
  pottery_count %>%
  mutate(phase = as.factor(phase)) %>% # seems need this line first
  inner_join(AD_zone_tidy) %>% 
  filter(!is.na(phase)) %>% 
  mutate(phase = case_when(
    phase %in% "pre-e"~ "Before European Contact",
    phase %in% "post-e" ~ "European Presence",
    phase %in% "ch-con" ~ "Chinese Presence")) %>% 
  mutate(phase = factor(phase,
                        level = c("Before European Contact",
                                  "European Presence",
                                  "Chinese Presence")))
# plot spatial pattern for pottery
plot_spatial_distribution_pottery_by_phase <- 
  ggplot() +
  geom_sf(data = AD_zone_tidy, fill = NA) +
  geom_sf(data = pottery_count_join, 
          aes(geometry = geometry, 
              fill =n), 
          alpha = 0.9)+
  facet_wrap(~ phase) +
  labs(fill = "frequency") +
  scale_fill_viridis(direction = -1) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        legend.key.size =  unit(3, "mm"))

plot_spatial_distribution_pottery_by_phase

ggsave(plot = plot_spatial_distribution_pottery_by_phase,
       here("analysis", "figures",
       "plot-spatial-distribution-pottery-by-period.jpg"),
       width = 135, 
       height = 31,
       dpi = 300,
       units = "mm")
```

@@

The scanned pottery drawings were acquired from Bureau of Cultural Affairs in Yilan. All drawing presents two-dimensional view of the section of a vessel with indications of metric measurements. The scanned drawings were imported into the Inkscape software for outlines tracing to remove additional information such as marks, lines, and measurements on the original drawings in order to obtain clear outline. Each traced half cross-section image was duplicated, flipped, and then joined with another one to create a 2D closed outline. Geometric morphometric analyses were conducted in the R software (www.rproject.org, Core-Team, 2015) using the functions included in the Momocs, a R package intended to quantify the shape and compare its variation, especially for outline analysis [@Bonhomme2014]. The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts. The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction.

>>>>

# Archaeological background and materials

Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from the central, undisturbed area of archaeological excavations at Kiwulan (Figure \@ref(fig:KWL-map); Figure \@ref(fig:KWL-sam-area)). Kiwulan is situated on a hill near a riverside at the northern margin of Yilan Province, which is characterized by a triangular alluvial plain facing eastwards the Pacific with high mountains on three other sides. Yilan is an ideal context to study peripheral colonial influence because Indigenous communities there were isolated by natural barriers that limited the frequency of direct contact with the Spanish and the Dutch settlers in northern Taiwan. The chronology of the archaeological deposits at Kiwulan consists of two cultural components, the upper component and the lower component, with a sterile layer in between [@Chen2007]. This paper focuses on the upper component, dated from AD 1350 to AD 1950, covering the late Iron Age and the historical period, which we define as the European presence in Taiwan in the early 17th century. The Dutch first occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626 <!-- add citation for these dates -->. In 1642, the Spanish was expelled by the Dutch in northern Taiwan, who took over their forts at Helping Dau in Keelung, and in Tamsui. Since then, western Taiwan was mostly under Dutch colonial rule until 1662 when the Kingdom of Tungning in Taiwan was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

The archaeological record of Kiwulan's upper component shows traces of foreign contacts, including Europeans in the 17th century, and waves of Chinese immigrants in the 19th century. Imported ceramics from mainland China, stoneware, and ornaments such as beads have been found in the upper component, indicating frequent long-distance trade activities in the 17th century during the European presence in northern Taiwan. In addition to artifacts, archaeological features such as burials, middens, and post-holes with _in-situ_ posts are widespread across  the 1-2 m thick deposit of the upper component, and demonstrate that Kiwulan was a continuously occupied large settlement site [@Chen2007]. To compare different foreign influences, we classified the upper component into three chronological phases: the pre-European, European, and Chinese periods. We identified these phases using 32 previously published radiocarbon ages <!-- add citaion to report with dates -->, excavation depth measurements, stratigraphic details reported by the excavators (colour, disturbance, etc.), and finds of chronologically diagnostic artifacts, such as blue and white porcelains, light grey glazed jars, and large dark brown glazed stoneware jars commonly used in the 17th century, and bricks and tiles used by the Chinese in the 19th century [@Hsieh2009; @Wang2011]. The deposit shows signs of continuous human occupation during each of the three phases across the sampling area.

The most abundant artefacts in the upper component are locally made ceramics, which are distributed throughout the sequence and across the site. More than 550,000 sherds were excavated, and around 1,200 vessels could be reconstructed. There are only two forms of local vessels, a cooking pot and a steamer made of two cooking pots stacked together with a clay filter layer in between. Those pots forms show high consistency in shape. They have a globular body with a short neck and wide mouth. The exterior surface below the neck is decorated with a wide variety of impressed geometric motifs. These vessels were probably used for cooking, with evidence of charred residues and carbon deposits frequently observed on vessel interiors and soot on their exteriors. Moreover, a lack of evidence of other utilitarian earthenware ceramics suggests these globular pots were mainly used for daily cooking. They are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Finger impressions and seams usually on the interior indicate they were pinched using slabs of clay and shaped by hand. This kind of pot has been widely found at archaeological sites during the late Iron Age and the historical period throughout the Yilan Plain.

@@

Petrographic analysis for 34 thin sections shows high percentage of inclusions (15-50%), including argillite (15-40%), metasandstone (1-10%),  sandstone fragments (1-6%), quartz (1-5%), and trace amounts of feldspar and slate. The size of the particles ranges from 500 micron to 1300 micron. In general, the globular vessel farbric presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz. The composition matches the mineralogical composition of local raw materials in the Yilan Plain [@Chen2016]. There is no substantial changes in the inclusions over time, indicating continuity in pottery fabric composition across the three periods.

@@

A total of `r total_no` largely intact vessels from pre-contact contexts (n = 32), post-European contexts (n = 27), and Chinese contact contexts (n = 14) were selected for Ellipic Fourier analysis. Most of them cover every vessel component from rim to bottom. To fully explore the pottery standardization, we also take the metric measurements into account for comparison. The total of 362 reconstructed pots (pre-European = 153, European = 173, Chinese = 36) were measured. The metric attributes include thickness of rim, neck, and body, and diameter of rim, neck, and body. For each pot, two measurements for thickness and diameter were took to compute an average value. To obtain information about the pottery shape, we also compare the ratio of Rim/Body thickness, and the ratio of Rim/Body diameter.

@@

We used scanned pottery drawings acquired from Bureau of Cultural Affairs in Yilan. All drawings present a two-dimensional view of vessel cross-sections with indications of metric measurements. The scanned drawings were imported into Inkscape (http://inkscape.org) where outlines were traced manually. Where only half cross-section images were available, these were duplicated, flipped, and then joined with another one to create a 2D closed outline for each vessel. Geometric morphometric analyses were conducted using the R software [@RCoreTeam] using the functions included in the Momocs, a R package intended to quantify and analyse shapes, especially for outline analysis [@Bonhomme2014]. The digitised outlines were converted into a list of successive x-y pixel coordinates for elliptic Fourier analysis (EFA), which assesses morphological differences among pottery shapes from three occupation contexts. The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction.
